---
title: "STAT- S 650 TIME SERIES ANALYSIS FINAL PROJECT- BITCOIN PRICE PREDICTION USING ARIMA AND SARIMA"
author: "VISWA SUHAAS PENUGONDA"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}

```

## Needed Libraries:

```{r}
library(tseries)
library(forecast)
library(ggplot2)
library(zoo)
library(tseries)
library(quantmod)
```

## Description of the Dataset:

The dataset is a comprehensive record of the daily trading activity of Bitcoin against the US dollar.
The price of Bitcoin for 3,145 Days is given in this Dataset, Starting from 17th September 2014 to 27th April 2023.

Each row of the dataset corresponds to a single day and includes the date, the opening price, the highest price, the lowest price, the closing price, the adjusted close price is adjusted for any corporate actions, such as stock splits, dividends, and rights offerings, which may affect the closing price of Bitcoin, and the trading volume for that day.
The prices are quoted in US dollars, and the volume is given in terms of the number of Bitcoin traded during that day.

The dataset can be used to track the price trends of Bitcoin to help traders and investors make informed decisions about their investments.
It is a valuable resource for those interested in the cryptocurrency market as it allows for the analysis of the cryptocurrency's performance over time.

## Load the Dataset

```{r}
BTC= read.csv("/Users/suhaaspenugonda/Downloads/BTC-USD.csv")
head(BTC)
tail(BTC)
dim(BTC)
summary(BTC)
```

## Variables in the Dataset:

1.  **Date:** The date of the trading day, in the format "YYYY-MM-DD".

2.  **Open:** The opening price of Bitcoin at the start of the trading day, quoted in US dollars.

3.  **High:** The highest price of Bitcoin reached during the trading day, quoted in US dollars.

4.  **Low:** The lowest price of Bitcoin reached during the trading day, quoted in US dollars.

5.  **Close:** The closing price of Bitcoin at the end of the trading day, quoted in US dollars.

6.  **Adj Close:** The adjusted closing price of Bitcoin at the end of the trading day, which takes into account any corporate actions or other events that may affect the price of Bitcoin.
    This price is also quoted in US dollars.

7.  **Volume:** The trading volume of Bitcoin during the trading day, measured in terms of the number of Bitcoin that were traded.

```{r}
Open <- as.numeric(BTC$Close)
#sum(is.na(Open))
Close <- as.numeric(BTC$Open)
#sum(is.na(Close))
High <- as.numeric(BTC$High)
#sum(is.na(High))
Low <- as.numeric(BTC$Low)
#sum(is.na(Low))
Volume <- as.numeric(BTC$Volume)
#sum(is.na(Volume))
```

## Hypothesis:

-   **Null Hypothesis:** There is no significant relationship between the closing price of Bitcoin and its future prices

-   **Alternate Hypothesis:** There is a significant relationship between the closing price and future prices of Bitcoin

## Check for Null values in the columns of the Dataset

```{r}
sum(is.na(Open))
sum(is.na(Close))
sum(is.na(High))
sum(is.na(Low))
sum(is.na(Volume))
```

## Remove the Null values in the columns of the Dataset

```{r}
Open <- na.omit(Open)
Close <- na.omit(Close)
High <- na.omit(High)
Low <- na.omit(Low)
Volume <- na.omit(Volume)
```

## Plots of the Time Series Data:

```{r}
plot(ts(Open),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Opening Prices")
plot(ts(Close),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Closing Prices")
plot(ts(High),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical High Prices")
plot(ts(Low),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Low Prices")
plot(ts(Adj_Close),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Adjusted Close Prices")
plot(ts(Volume),type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Volumes")
```

## Smoothing the above Time Series Plots

```{r}
ws = 50
Open_s = filter(Open, rep(1/ws, ws), sides = 2)
plot(Open_s,type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Opening Prices")

ws = 50
Close_s = filter(Close, rep(1/ws, ws), sides = 2)
plot(Close_s,type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Closing Prices")

ws = 50
High_s = filter(High, rep(1/ws, ws), sides = 2)
plot(High_s,type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical High Prices")

ws = 50
Low_s = filter(Low, rep(1/ws, ws), sides = 2)
plot(Low_s,type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Low Prices")

ws = 200
Volume_s = filter(Volume, rep(1/ws, ws), sides = 2)
plot(Volume_s,type="l", xlab="Time", ylab = "Price (USD)", main="Bitcoin Historical Volumes")
```

## Differencing the variables

```{r}
diff_Open <- diff(Open)
diff_Close <- diff(Open)
diff_High <- diff(High)
diff_Low <- diff(Low)
diff_Volume <- diff(Volume)
```

## Augmented Dickey-Fuller (ADF)- Stationality check

```{r}
Open_result <- adf.test(diff_Open)
cat("ADF test results for Open: ", Open_result$p.value, "\n")
Close_result <- adf.test(diff_Close)
cat("ADF test results for Close: ", Close_result$p.value, "\n")
High_result <- adf.test(diff_High)
cat("ADF test results for High: ", High_result$p.value, "\n")
Low_result <- adf.test(diff_Low)
cat("ADF test results for Low: ", Low_result$p.value, "\n")
Volume_result <- adf.test(diff_Volume)
cat("ADF test results for Volume: ", Volume_result$p.value, "\n")
```

Since the p values for all the variables is less than 0.05, we can say that the variables are stationary.

## Computation of the Autocorrelation functions for all the variables

```{r}
acf1 <- acf(diff_Open, main = "ACF for Residuals of Open")
acf2 <- acf(diff_Close, main = "ACF for Residuals of Close")
acf3 <- acf(diff_High, main = "ACF for Residuals of High")
acf4 <- acf(diff_Low, main = "ACF for Residuals of Low")
```

## Computation of the cross-correlation function for the residuals of the Open/Close, High/Low variables

```{r}
ccf_res1 <- ccf(diff_Open, diff_Close, main = "Cross-Correlation of Residuals of Open/Close")
ccf_res2 <- ccf(diff_High, diff_Low, main = "Cross-Correlation of Residuals of High/Low")
```

## Spectral analysis of each series

```{r}
Open_Spec <- spec.pgram(diff_Open,spans=c(50,50), plot = TRUE, , main="Spectral Distribution of Bitcoin Historical Opening Prices")
Close_Spec <- spec.pgram(diff_Close,spans=c(50,50), plot = TRUE, main="Spectral Distribution of Bitcoin Historical Closing Prices")
High_Spec <- spec.pgram(diff_High,spans=c(50,50), plot = TRUE, main="Spectral Distribution of Bitcoin Historical High Prices")
Low_Spec <- spec.pgram(diff_Low,spans=c(50,50), plot = TRUE, main="Spectral Distribution of Bitcoin Historical Low Prices")
```

In the above plots, Window sizes 50 is used to plot the Estimated Spectral Density of the 'Open', 'Close', 'High' and 'Low' measures while 'Volume' had a window size of 100.
50 gave us the correct level of smoothing.
The window size of 50 also preserve the variance in the data.
Most of the data in the plot suggest that the Bitcoin price and frequency are inversely proportional to each other.

## Analysis by Fitting linear dynamic models

```{r}
N = length(diff_Close)

A = c(0,1)
B = c(0,0,1)
C = c(0,0,0,1)

Open_1 =  filter(diff_Open,A,sides=1)
Open_2 =  filter(diff_Open,B,sides=1)
Open_3 =  filter(diff_Open,C,sides=1)
Close_1 =  filter(diff_Close,A,sides=1)
Close_2 =  filter(diff_Close,B,sides=1)
Close_3 =  filter(diff_Close,C,sides=1)
High_1 =  filter(diff_High,A,sides=1)
High_2 = filter(diff_High,B,sides=1)
High_3 = filter(diff_High,C,sides=1)
Low_1 =  filter(diff_Low,A,sides=1)
Low_2 =  filter(diff_Low,B,sides=1)
Low_3 =  filter(diff_Low,C,sides=1)
Volume_1 = filter(diff_Volume,A,sides=1)
Volume_2 = filter(diff_Volume,B,sides=1)
Volume_3 = filter(diff_Volume,C,sides=1)
```

### Model 1

Y(t) = x1(t-1)+x1(t-2)+x1(t-3)

```{r}
mod_1 = lm(diff_Close[3:length(diff_Close)]~Open_1[3:length(Open_1)]+Open_2[3:length(Open_2)])
summary(mod_1)
AIC(mod_1)
BIC(mod_1)
```

### Model 2

Y(t) = x1(t-1)+x1(t-2)+x2(t-1)+x2(t-2)

```{r}
mod_2 = lm(diff_Close[3:length(diff_Close)]~Open_1[3:length(Open_1)]+Open_2[3:length(Open_2)]+High_1[3:length(High_1)]+High_2[3:length(High_2)])
summary(mod_2)
AIC(mod_2)
BIC(mod_2)
```

### Model 3

Y(t) = x1(t-1)+x1(t-2)+x2(t-1)+x2(t-2)+x3(t-1)+x3(t-2)

```{r}
mod_3 = lm(diff_Close[3:length(diff_Close)]~Open_1[3:length(Open_1)]+Open_2[3:length(Open_2)]+High_1[3:length(High_1)]+High_2[3:length(High_2)]+Low_1[3:length(Low_1)]+Low_2[3:length(Low_2)])
summary(mod_3)
AIC(mod_3)
BIC(mod_3)
```

### Inference from the above Information

Looking at the scores of the three models, we can see the AIC and BIC scores of model 3 are less compared to model 2 with model 1 giving the highest scores , hence model 3 is chosen.

```{r}
mod_1.resid = mod_1$residuals
plot(mod_1.resid)

acf(mod_1.resid, 10, main="output")
Box.test(mod_1.resid,2)

spectrum(mod_1.resid,log="no")
H = spectrum(mod_1.resid,spans=c(20,20))

ypred2=predict(mod_1)
plot(diff_Close[3:N],ypred2,xlab = 'observed', ylab = 'predicted')

```

```{r}
Open_a <- diff_Open-mean(diff_Open)
Open_a = ts(Open_a,start=1,end=N)
Close_a <- diff_Close-mean(diff_Close)
Close_a = ts(Close_a,start=1,end=N)
High_a <- diff_High-mean(diff_High)
High_a = ts(High_a,start=1,end=N)
Low_a <- diff_Low-mean(diff_Low)
Low_a = ts(Low_a,start=1,end=N)
Vol_a <- diff_Vol-mean(diff_Vol)
Vol_a = ts(Vol_a, start=1, end=N)

Open_a1 = lag(Open_a,-1)
Close_a1 = lag(Close_a,-1)
High_a1 = lag(High_a,-1)
Low_a1 = lag(Low_a,-1)

Open_a2 = lag(Open_a1,-1)
Close_a2 = lag(Close_a1,-1)
High_a2 = lag(High_a1,-1)
Low_a2 = lag(Low_a1,-1)

N = length(Close_a)
N1 = N - 10
N2 = (N1+1):N

X1 = cbind(Open_a1, Close_a1, High_a1, Low_a1)
X2 = cbind(Open_a1, Open_a2, High_a1, High_a1, Low_a1, Low_a2)

Closef = Close_a[N2]
X1f = X1[N2,]
X2f = X2[N2,]

Close_a = Close_a[1:N1]
X1 = X1[1:N1,]
X2 = X2[1:N1,]
```

### Model 1
```{r}
mod_1a = arima(Close_a, xreg=X1, order=c(0,0,1))
mod_1a

AIC(mod_1a)
BIC(mod_1a)
```


### Model 2
```{r}
mod_2a = arima(Close_a, xreg=X2, order=c(1,0,0))
mod_2a

AIC(mod_2a)
BIC(mod_2a)
```

### Model 3
```{r}
mod_3a = arima(Close_a, order = c(3,0,3))
mod_3a

AIC(mod_3a)
BIC(mod_3a)
```

### Inference from the above Information

Looking at the scores of the three models, we can see the AIC and BIC scores of model 1 are less compared to model 2 with model 3 giving the highest scores , hence model 1 is chosen.

## Examine the residuals of the model using Autocovariance and Spectral Analysis
```{r}
cRes1 = residuals(mod_1a)
plot(cRes1)

acf(cRes1[3:N1], 10, main="output")

Box.test(cRes1,lag=5)
spectrum(cRes1[3:N1],log='no')
```

## Plots of the predictions of the model using Hypothetical inputs

```{r}
ns=10
yc=predict(mod_1a,newxreg=X1f[1:ns,])
yclose = yc$pred
ts.plot(Closef[1:ns], yclose, lty=1:2)

yc=predict(mod_2a,newxreg=X2f[1:ns,])
yclose = yc$pred
ts.plot(Closef[1:ns], yclose, lty=1:2)

yc=predict(mod_3a,n.ahead=ns)
yclose = yc$pred
ts.plot(Closef[1:ns], yclose, lty=1:2)
```

On observing the above graph, we can say that the models 1& 2 are good at making predictions as the predicted values are closer to the actual values, with model 3 making good predictions at the beginning, trailing off later.










